Schema Change: Challenge Problems

### 2024-05-08 paper.tex
Related Work

% define natbib citet
\newcommand{\citet}[1]{\citeauthor*{#1} \cite{#1}}

@@

@@

~
\textcolor{red}{[JE] We could expand our context to include conventional DB systems not just live/local. It is a widespread problem, attested by the many tools to assiste with it. We could include in our problems a description of a typical solution with a standard DB as a sort of baseline. }

@@

>>>>

% define natbib citet
\newcommand{\citet}[1]{\citeauthor*{#1}~\cite{#1}}

@@

\textcolor{red}{[JE] We should adopt the term Schema Evolution, which is favored in the research literature, defined to include schema change, data migration, and query rewriting. }

@@

@@

\section{Related Work}

Schema evolution has long been a major problem for SQL databases, which provide suprisingly little help. SQLite~\cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes. MYSql~\cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards. In \textit{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

\citet{bernstein07} observed in 2007 ``There are hardly any schema evolution tools today. This is rather surprising since there is a huge literature on schema evolution spanning more than two decades.'' There are now more tools available. Some tools such as Liquibase~\cite{liquibase} and PlanetScale\cite{planetscale} could be characterized as version control and continuous integration/deployment for schema. They track schema changes and can calculate diffs in the form of SQL DDL statements to convert one schema version to another, but do not help migrate data. Unfortunately comparing schemas can be ambiguous about the intention of changes. For example has a column been renamed or has it been deleted and a new one created? That distinction makes a big difference to the data in that column. EvolveDB\cite{evolvedb} addresses this ambiguity by reverse-engineering the schema into a richer data model and tracking the edits to that model within an IDE. This more precise edit history can be used to infer higher level intentions of a schema change, which then generate SQL scripts to evolve the database. EdgeDB~\cite{edgedb} resolves ambiguities by asking questions of the developer, with some answers supplying custom migration code in a proprietary query language.

Some tools provide a Domain Specific Language (DSL) to describe schema evolution. Rails Migrations~\cite{RailsMigrations} embeds a DSL in Ruby to describe a range of common schema evolutions. However that falls short of the range of evolutions described by \citet{ambler06} which require the addition of custom Ruby or SQL code. \citet{curino08} initiated a large thread of research on Database Evolution Languages (DEL) by defining a Schema Modification Operator (SMO) as ``a function that receives as input a relational schema and the underlying database, and produces as output a (modified) version of the input schema and a migrated version of the database''. SMOs can also rewrite queries to accomodate schema changes. \citet{herrmann15} defined a relationally complete DEL and then extended it into a Bidirectional Database Evolution Language (BiDEL)~\cite{herrmann17} that appears capable of handling our Extract Entity and Multiplicity Change challenges, including divergence.

Schema evolution is also a problem for NoSQL~\cite{sadalage12} databases. While such databases are sometimes called ``schemaless'' in effect that means the schema is left implicit and tools must try to infer it~\cite{storl20, storl22}. \citet{scherzinger13} define a set of operators like the relational SMOs discussed above. \citeauthor*{chillon21}~\cite{chillon21, chillon22} offer a more comprehensive set of SMOs that appear capable of handling the data migration aspects of our Extract Entity and Multiplicity Change challenges. \citet{Cambria} uses lenses~\cite{Foster2007} for bidirectional transformation of JSON. None of the approaches to NoSQL evolution have yet extended like their SQL cousins into updating code or rewriting queries.

Schema evolution has also been studied for Object Oriented Databases(OODB)~\cite{li99,banerjee87}. Smalltalk~\cite{Goldberg80} is itself an OODB, persisting all object instances in an ``image file'' with limited evolution capabilities incorporated in the programming environment~\cite[pp.252-272]{Goldberg80}. Gemstone turns the Smalltalk image into a production-quality database and accordingly provides a sophisticated schema evolution API~\cite{Gemstone}.

\begin{comment}

dataframe diff
https://homepages.inf.ed.ac.uk/csutton/publications/datadiff.pdf

Durable computing: pins runtime data frames to code versions. Doesn’t handle problem of global data schema change.
https://restate.dev/blog/solving-durable-executions-immutability-problem/
https://restate.dev/blog/code-that-sleeps-for-a-month/

Tijs on Live programming and model-driven engineering
https://ir.cwi.nl/pub/23770
https://dl.acm.org/doi/10.1007/s10270-017-0608-7
https://dl.acm.org/doi/10.1145/3276604.3276611
https://ir.cwi.nl/pub/21858

From Geoffrey:
1) Finding Data Compatibility Bugs with JSON Subschema Checking
https://dl.acm.org/doi/pdf/10.1145/3460319.3464796
I just found this paper this morning so I haven't read it yet, but the abstract seems very relevant. It's about automatically determining whether two JSON schemas are "compatible", with a subtyping metaphor.
2) Confluent Schema Registry:
https://docs.confluent.io/platform/current/schema-registry/index.html
An interesting example of how schema change is handled in the wild. In the Cambria essay we wrote some reflections on it which basically capture my opinions.
\end{comment}
